\documentclass{article}
\input ../preamble
\parindent = 0em

%\newcommand{\solution}[1]{}
\newcommand{\solution}[1]{\bigskip {\color{red} {\bf Solution}: #1}}

\begin{document}


\centerline{\bf TTIC 31230 Fundamentals of Deep Learning, winter 2019}
\medskip
\centerline{\bf Quiz 2}

\bigskip
{\bf Problem 1. 25 points.} Equations defining a UGRNN are given below.


\begin{eqnarray*}
\tilde{R}_t[b,j] & = & \left(\sum_i\;W^{h,R}[j,i]{\color{red} h_{t-1}[b,i]}\right) + \left(\sum_k W^{x,R}[j,k]{\color{red} x_t[b,k]}\right) - B^R[j] \\
\\
R_t[b,j] & = & \mathrm{\tanh}(\tilde{R}_t[b,j]) \\
\\
\tilde{G}_t[b,j] & = & \left(\sum_i\;W^{h,G}[j,i]{\color{red} h_{t-1}[b,i]}\right) + \left(\sum_k W^{x,G}[j,k]{\color{red} x_t[b,k]}\right) - B^G[j] \\
\\
G_t[b,j] & = & \mathrm{\sigma}(\tilde{G}_t[b,j]) \\
\\
{\color{red} h_t[b,j]} & = & G_t[b,j]{\color{red} h_{t-1}[b,j]} + (1-G_t[b,j])R_t[b,j]
\end{eqnarray*}

(a) Rewrite the first equation defining $\tilde{R}_t$ using $\pluseq$ loops instead of summations assuming that all computed tensors are initialized to zero.


\solution{
\begin{eqnarray*}
\mbox{for}\; b,j,i\;\tilde{R}_t[b,j] & \pluseq & W^{h,R}[j,i]h_{t-1}[b,i] \\
\\
\mbox{for}\;b,j,k\;\tilde{R}_t[b,j] & \pluseq & W^{X,R}[k,i]x_t[b,k] \\
\\
\mbox{for}\;b,j\;\tilde{R}_t[b,j] & \minuseq & B^R[j]
\end{eqnarray*}
}

\medskip
(b) Give $\pluseq$ loops for the backward computation for your solution to part (a) using the convention that
parameter gradients are averaged over the batch and where the batch size is $B$.

\solution{
\begin{eqnarray*}
\mbox{for}\; b,j,i\;W^{h,R}.\grad[j,i] & \pluseq & \frac{1}{B}\;h_{t-1}[b,i]\tilde{R}_t.\grad[b,j] \\
\\
\mbox{for}\; b,j,i\; h_{t-1}.\grad[b,j] & \pluseq & W^{h,R}[j,i]\tilde{R}_t.\grad[b,j] \\
\\
\mbox{for}\; b,j,k\;W^{x,R}.\grad[j,k] & \pluseq & \frac{1}{B}\;x[b,k]\tilde{R}_t.\grad[b,j] \\
\\
\mbox{for}\; b,j,k\; h_{t-1}.\grad[b,j] & \pluseq & W^{x,R}[j,k]\tilde{R}_t.\grad[b,j] \\
\\
\mbox{for}\;b,j\;B^R.\grad[j] & \minuseq & \frac{1}{B}\;\tilde{R}_t.\grad[b,j]
\end{eqnarray*}
}

\bigskip
{\bf Problem 2. 25 points.} Images have translation invariance --- a person detector must look for people at various places in the image.  Translation invariance is the motivation for
convolution --- all places in the image are treated the same.

\medskip
Images also have some degree of scale invariance --- a person detector must look for people of different sizes
(near the camera or far from the camera).  We would like to design a deep architecture that treats all scales (sizes) the same in a manner that similar to the way CNNs
treat all places the same.

\medskip
Consider a batch of images
$I[b,x,y,c]$ where $c$ ranges over the three color values red, green, blue. We start by constructing an ``image pyramid'' $I_s[x,y,c]$.
We assume that the original image $I[b,x,y,c]$ has spatial dimensions $2^k$ and construct images $I_s[b,x,y,c]$ with spatial dimensions $2^{k-s}$
for $0 \leq s \leq s_{\mathrm{max}} < k$.  These images are defined by the following equations.
\begin{eqnarray*}
  I_0[b,x,y,c] & = & I[b,x,y,c] \\
  \\
  I_{s+1}[b,x,y,c] & = & \frac{1}{4}\left(\begin{array}{l} I_s[b,2x,2y,c] + I_s[b,2x+1,2y,c] \\ + I_s[b,2x,2y+1,c] + I_s[b,2x+1,2y+1,c]\end{array}\right)
\end{eqnarray*}


We want to compute a set of layers $L_{s,\ell}[b,x,y,i]$ where $s$ is the scale and $\ell$ is the level of processing.  First we set
$$L_{0,s}[b,x,y,c] = I_s[b,x,y,c].$$
The layers $L_{\ell,0}[b,x,y,i]$ can be computed using the standard CNN equations holding the scale at zero.

\medskip
Give an equation for a linear threshold unit to compute $L_{\ell+1,s+1}[b,x,y,j]$ from $L_{\ell,s+1}[b,x,y,j]$ {\bf and} $L_{\ell+1,s}[b,x,y,j]$.  Assume that the
spatial dimension of $L_{\ell,s}$ is $2^{k-s}$ and use an appropriate stride between $L_{\ell+1,s+1}[b,x,y,j]$ and $L_{\ell+1,s}[b,x,y,j]$.  Use parameters
$W_{\ell+1,\rightarrow}[\Delta x, \Delta y, i,j]$ for the dependence of $L_{\ell+1,s}$ on $L_{\ell,s}$ and
parameters
$W_{\ell+1,\uparrow}[\Delta x, \Delta y, i,j]$ for the dependence of $L_{\ell+1,s+1}$ on $L_{\ell+1,s}$.  Use $B_{\ell+1}[j]$ for the threshold.
Note that these parameters do not depend on $s$ --- they are scale invariant.

\solution{
  $$L_{\ell+1,s+1}[b,x,y,j] = \sigma\left(
  \begin{array}{l}
    \sum_{\Delta x,\Delta y,i}\;W_{\ell+1,\rightarrow}[\Delta x, \Delta y, i,j]L_{\ell,s+1}[b,x+\Delta x,\;y+\Delta y,\;i] \\
    \sum_{\Delta x,\Delta y,i}\;W_{\ell+1,\uparrow}[\Delta x, \Delta y, i,j]L_{\ell+1,s}[b,2x+\Delta x,\;2y+\Delta y,\;i] \\
    - B_{\ell+1}[j]\end{array}\right)$$
}

  

\medskip
{\bf Problem 3. 25 points.}

Modify the equations for a UGRNN from problem 1 to form a data-dependent data-flow CNN for vision --- an Update-Gate CNN (UGCNN).
More specifically, give equations analogous to those for UGRNN for computing a CNN ``box'' $L_{\ell+1}[b,x,y,j]$ from $L_\ell[b,x,y,i]$ (stride 1) using a computed ``gate box'' $G_{\ell+1}[b,x,y,j]$ and an ``update box'' $R_{\ell+1}[b,x,y,j]$.
$$\Phi = (W^{L,R}_{\ell+1}[\Delta x, \Delta y,j,j'], B^R_{\ell+1}[j], W^{L,G}_{\ell+1}[\Delta x, \Delta y,j,j'],\;B^G_{\ell+1}[j])$$

\solution{
  \begin{eqnarray*}
    R_{\ell+1}[b,x,y,j] & = & \mathrm{tanh}\left(\left(\sum_{\Delta x,\Delta y,j'}\;W^{L,R}_{\ell+1}[\Delta x, \Delta y, j',j]
    L_{\ell}[b,x+\Delta x,\;y+\Delta y,\;j']\right) - B^R_{\ell+1}[j]\right) \\
    \\
    G_{\ell+1}[b,x,y,j] & = & \sigma\left(\left(\sum_{\Delta x,\Delta y,i}\;W^{L,G}_{\ell+1}[\Delta x, \Delta y, i,j']
    L_{\ell}[b,x+\Delta x,\;y+\Delta y,\;j']\right) - B^G_{\ell+1}[j]\right) \\
    \\
    L_{\ell+1}[b,x,y,j] & = & G_{\ell+1}[b,x,y,j]L_\ell[b,x,y,j] + (1-G_t[b,x,y,j])R_t[b,x,y,j]
  \end{eqnarray*}
}

{\bf Problem 4. 25 points.}
This problem is on CNNs for sentences.  We consider a model with parameters
$$\Phi = (e[w,i], W_1[\Delta t,i,i'], B_1[i],\ldots,W_L[\Delta t,i,i'],B_L[i])$$
The matrix $e$ is the word embedding matrix where $e[w,I]$ is the vector embedding of word $w$.

\medskip
(a) Give an equation for the convolution layer $L_0[b,t,i]$ as a function of the word embeddings and the input sentence $w_1,\ldots,w_T$.

\solution{
  $$L_0[b,t,i] = e[w[t],i]$$}

\medskip
(b) Give an equation for $L_{\ell+1}[b,t,i]$ as a function of $L_\ell[b,t,i]$ and the parameters $W_{\ell+1}[\Delta t,i',i]$ and $B_{\ell+1}[i]$
and where $L_{\ell+1}$ is computed stride 2.

\solution{
$$L_{\ell+1}[b,t,i] = \sigma\left(\left(\sum_{\Delta t,i'}\;W_{\ell+1}[\Delta t,i',i]L_\ell[2t + \Delta t,i']\right) - B_{\ell+1}[i]\right)$$
}
  
\medskip
(c) Assuming all computations can be done in parallel as soon the inputs have been computed, what is the {\bf parallel} order
of run time for this convolutional model as a function of the input length $T$ and the number of layers $L$ (assume all parameter tensors
of size $O(1)$). Compare this with the parallel run time of an RNN.

\solution{The CNN has $O(L)$ parallel run time while the RNN is $O(T)$ or $O(T+L)$ with $L$ layers of RNN.}


\end{document}
