\input ../../SlidePreamble
\input ../../preamble

\begin{document}

{\Huge

  \centerline{\bf TTIC 31230, Fundamentals of Deep Learning}
  \bigskip
  \centerline{David McAllester, Autumn 2020}

\vfill
  \centerline{\bf AGI: Logic}
  \vfill
  \vfill

\slide{Logic Remains Central to Computer Science}

Compilers remain essential to computing.  Compilers have a precise (logical) notion of correctness.

\vfill
The correctness of compilers underlies the correctness of the programs they compile.

\vfill
I.J. Good's intelligence explosion presumably requires AI systems to be able to write (logically) correct programs.

\slide{Logic in Common Sense}

Certain facts are obvious.

\vfill
A king on empty chess board can reach every square (obvious).

\vfill
A knight on an empty chess board can reach every square (true but not obvious).

\slide{Logic in Common Sense}

Consider a graph with colored nodes.

\vfill
If every edge is between nodes of the same color, then any path connects nodes of the same color.

\vfill
Consider a swiss chocolate bar of $3 \times 5$ little squares.

\vfill
How many breaks does it take to reduce this to fifteen unconnected squares?

\vfill
What inference happens when one observes that each break increases the number of pieces by one?

\slide{Logic in Lexical Semantics}

Let $e$ range over ``events''.

\vfill
$e:\mathrm{give}(a_1,x,a_2) \Rightarrow \mathrm{had}(a_1,x,\mathrm{before}(e)) \;\wedge\; \mathrm{had}(a_2,x,\mathrm{after}(e))$

\vfill
This is related to Davidsonian semantics for natural language (1969) and the situation calculus of McCarthy and Hayes (1968).

\slide{Logic for Computation}

Bottom-up logic programming is distinguished by its relationship to dynamic programming algorithms.

\vfill
$$\mathrm{At}(x) \Rightarrow \mathrm{Reachable}(x)$$

\vfill
$$\mathrm{Reachable}(x) \wedge \mathrm{CanGo}(x,y) \Rightarrow \mathrm{Reachable}(y)$$

\vfill
This defines a linear time algorithm for reachability.

\slide{Logic for Computation}

A set of inference rules (horn clauses) in which each term in the conclusion appears in some antecedent is called a local logic program.

\vfill
It can be shown that local logic programs ``captures the complexity class $P$'' --- they can express {\bf all and only} polynomial time decidable relations.

\slide{The MathZero Program}

Is it possible to achieve super-human performance in mathematics through self-play?

\slide{Autoformalization}

Christian Szegedy (adversarial images, batch normalization,  the inception network) has written a position paper
taking the position that translation of natural language mathematics into formal machine-verifiable statements will provide
a path to AGI.

\vfill
A Promising Path Towards Autoformalization and General Artificial Intelligence, Christian Szegedy, Intelligent Computer Mathematics, 2020.

\slide{END}

}
\end{document}
