\input ../../SlidePreamble
\input ../../preamble
\def\entity#1{\vfill {[#1] =~}}

\begin{document}

{\Huge

  \centerline{\bf TTIC 31230, Fundamentals of Deep Learning}
  \bigskip
  \centerline{David McAllester, Autumn 2020}

\vfill
\centerline{\bf AGI: Natural Language}
  \vfill
  \vfill

\slide{Natural Language for Knowledge Representation}

{\huge
President Donald Trump on Tuesday insisted he was serious when he % 11
revealed that he had directed his administration to slow coronavirus % 10
testing in the United States, shattering the defenses of senior White % 11
House aides who argued Trump’s remarks were made in jest. %10

\entity{the order} [Trump] directed his administration to reduce [{\large COVID} testing].

\entity{the revelation} [Trump] revealed [the order].

\entity{the claim} [the order] was in jest.

\entity{the assertion} [the aides] made [the claim].
}

\slide{Natural Language for Knowledge Representation}

{\huge
President Donald Trump on Tuesday insisted he was serious when he % 11
revealed that he had directed his administration to slow coronavirus % 10
testing in the United States, shattering the defenses of senior White % 11
House aides who argued Trump’s remarks were made in jest. %10

\entity{the aides} Senior White House aides

\entity{the defense} [the assertion] defended [Trump].

\entity{the insistence} on Tuesday [Trump] insisted [the order] was serious.

\entity{the shattering} [the insistence] shattered [the defense].
}

\slide{Natural Language for Knowledge Representation}

Is natural language a reflection of a general learning or knowledge representation architecture?

\vfill
Will some NLP-motivated memory-based Transformer-like architecture ultimately underly general intelligence?

\slide{Natural Language Semantics}

Thousands of civilians have fled advances by Syrian government forces in eastern Ghouta as ...

\slide{Stanford Parse Tree}
\begin{verbatim}
(NP (NP (NNS Thousands))
    (PP (IN of) (NP (NNS civilians))))
(VP (VBP have)
    (VP (VBN fled)
        (NP (NNS advances))
        (PP (IN by) (NP (NP (JJ Syrian)
                            (NN government)
                            (NNS forces))
                        (PP (IN in) (NP (JJ eastern)
                                        (NNP Ghouta))))))))
\end{verbatim}

\slideplain{Stanford Dependencies}

\begin{verbatim}
root(ROOT-0, fled-5)
  aux(fled-5, have-4)
  nsubj(fled-5, Thousands-1)
    nmod(Thousands-1, civilians-3)
      case(civilians-3, of-2)
  dobj(fled-5, advances-6)
  nmod(fled-5, forces-10)
    case(forces-10, by-7)
    amod(forces-10, Syrian-8)
    compound(forces-10, government-9)
    nmod(forces-10, Ghouta-13)
      case(Ghouta-13, in-11)
      amod(Ghouta-13, eastern-12)
\end{verbatim}

\slideplain{Just Parantheses}

\begin{verbatim}
(Thousands of civilians)
(have fled)
(advances (by (Syrian government forces))
          (in eastern Ghouta))
\end{verbatim}

\slideplain{Reference (Entity Linking)}

{\color{blue} Thousands of civilians have fled advances by Syrian government forces in eastern Ghouta as}
Damascus makes rapid gains against the last major rebel enclave near the capital.

\vfill
Damascus $\Rightarrow$ Assad

\vfill
Rapid Gains $\Rightarrow$ advances-6

\vfill
the last major rebel enclave ... $\Rightarrow$ Ghouta

\vfill
the capital $\Rightarrow$ Damascus

\slide{Reference vs. Composition}

Functional programming is compositional

\vfill
$$x = f(y,z)$$

\vfill
The meaning of $x$ is computed by $f$ from the meaning of $y$ and $z$.

\vfill
But in language we typically have that $f(y,z)$ is a mention and $x$ is its referent.

\vfill
(the last (major rebel enclave) (near (the capital)))

\vfill
$$x = (\mbox{the last $Q$ $P$})$$

\slide{END}

}
\end{document}
